Compositional Quantitative Resource Analysis of C Programs

### 2014-06-24 paper/main.tex
update to the semantics

For simplicity, we assume that local variables and global
variables are always disjoint sets. % This simplifies both
% the semantics and the description of our analysis.  
The mutable state $\state$ can be accessed and modified, the access is
written functionally and is defined by
$$
(\theta, \gamma)(x) =
\left\{
\begin{array}{lr}
\theta(x) & \mbox{ if } x \in \dom \theta \\
\gamma(x) & \mbox{ if } x \in \dom \gamma
\end{array}
\right..
$$
For a program state $\state$, we write $\state[x \mapsto v]$ for the
program state that maps $x$ to $v$ and behaves as $\state$ for all
other variables, regardless whether $x \in \dom \theta$ or $x \in \dom
\gamma$.  Similarly, we just write $\state(x)$ to obtain the value of
$x$ in program state $\state$.

The small-step semantics is standard, except that it tracks the
resource consumption of a program.  The semantics is oblivious to the
resource of interest for the user of our system.  We achieve this
independence by parameterizing evaluations with a resource
\emph{metric} $M$; a tuple of nine rational numbers, one map $M_e :
\text{Expr} \to \Q$ from expressions to rational numbers, and one map
$M_t : \Z \to \Q$ from integers to rational numbers.
$$
M = (
M_a,
M_b,
M_r,
M_u,
M_f,
M_l,
M_c^1,
M_c^2,
M_q,
M_e(E),
M_t(n)
)
$$
Each of these rational numbers indicates the amount of resource
consumed by a corresponding operation.  If one of these numbers is
negative, it means that resources are released by the program.  The
metric that we use in the implementation can also depend on other
information that is statically available such as the name of the called
function or the number of arguments.

@@

\pref{fig:opsem} contains the reduction rules of the semantics.
The rules define a rewrite system for program configurations of the
form $(\state, S, \cont, \cost)$, where $\state$ is the program state,
$S$ is the statement being executed, $\cont$ is a continuation that
describes what remains to be done after the execution of $S$, and
$\cost$ is the non-negative number of resources available for further
execution.  All rules that can decrease $\cost$ have the implicit side
condition that the resource quantity available after the step is
non-negative.

@@

The intuitive meaning of an evaluation judgement $(\state, S, \cont,
\cost) \smallstep{M}^* (\state', S', \cont', \cost')$ is the following.
If the statement $S$ is executed in program state $\state$, with
continuation $\cont$, and with $\cost$ resources available
then---after a finite number of steps---the evaluation will reach the
new machine state $(\state', S', \cont')$ and there are $\cost'$
resources available.  The resource consumption of the evaluation up to
this point is then $\cost - \cost'$.  If this difference is negative
then resource became available during the execution.

\pref{lem:sem1} summarizes the main properties of the resource
counter of the semantics.  It can be proved by induction on the number
of steps.
\begin{lemma}
\label{lem:sem1}
  \begin{enumerate}
  \item If $(\state, S, \cont, \cost) \smallstep{M}^* (\state', S',
    \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
    \smallstep{M}^* (\state', S', \cont', \cost' {+} n)$.
  \item If $(\state, S, \cont, \cost) \smallstep{M}^* (\state', S',
    \cont', \cost')$ and $(\state, S, \cont, d) \smallstep{M}^* (\state', S',
    \cont', d')$ then $\cost - \cost' = d - d'$.
  \end{enumerate}
\end{lemma}
%
If we assume that we have a \emph{positive resource metric}, that is,
a metric that assigns a positive cost to each evaluation step then we
can link bounds on the resource consumption under the metric to
termination.  This is formalized by \pref{lem:sem_term}.
%
\begin{lemma}
 \label{lem:sem_term}
 Let $M$ be a positive metric and assume that there exists $b \in
 \Qplus$ such that $(\state, S, \cont, \cost) \smallstep{M}^*
 (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.  Then
 $(\state, S, \cont, d) \smallstep{M}^* (\state'', \code{skip}, \Kstop,
 d')$ for some $d$, $d'$ and $\state''$.
\end{lemma}

>>>>

For simplicity, we assume that local variables and global variables
are always disjoint.  We just write $\state(x)$ to obtain the value of
$x$ in program state $\state$.  Such a lookup is defined as
$$
(\theta, \gamma)(x) =
\left\{
\begin{array}{lr}
\theta(x) & \mbox{ if } x \in \dom \theta \\
\gamma(x) & \mbox{ if } x \in \dom \gamma
\end{array}
\right..
$$
For a program state $\state$, we similarly write $\state[x \mapsto v]$
for the program state that maps $x$ to $v$ and behaves as $\state$ for
all other variables, regardless whether $x \in \dom \theta$ or $x \in
\dom \gamma$.

The small-step semantics is standard, except that it tracks the
resource consumption of a program.  The semantics is oblivious to the
resource of interest for the user of our system.  We achieve this
independence by parameterizing evaluations with a resource
\emph{metric} $M$; a tuple of nine rational numbers, one map $M_e :
\text{Expr} \to \Q$ from expressions to rational numbers, and one map
$M_t : \Z \to \Q$ from integers to rational numbers.
$$
M = (
M_a,
M_b,
M_r,
M_u,
M_f,
M_l,
M_c^1,
M_c^2,
M_q,
M_e(E),
M_t(n)
)
$$
Each of these rational numbers indicates the amount of resource
consumed by a corresponding operation.  If the assigned resource cost
is negative then it means that resources are released.  The metric
that we use in the implementation can also depend on other information
that is statically available such as the name of the called function
or the number of arguments.

@@

\pref{fig:opsem} contains the reduction rules of the semantics.  The
rules define a rewrite system for program configurations of the form
$(\state, S, \cont, \cost)$, where $\state$ is the program state, $S$
is the statement being executed, $\cont$ is a continuation that
describes what remains to be done after the execution of $S$, and
$\cost \in \Q$ is the non-negative number of resources available for
further execution.  All rules that can decrease $\cost$ have the
implicit side condition that the resource quantity available
\emph{before} the step is non-negative.  This means that we allow
$(\state, S, \cont, \cost)$ with $c<0$ on the right-hand side
transition relation $\smallstep{M}^*$ to indicate that the execution
ran out of reassesses.  However, every execution that reaches such a
state is stuck.

@@

The intuitive meaning of an evaluation judgement $(\state, S, \cont,
\cost) \smallstep{M}^* (\state', S', \cont', \cost')$ is the
following.  If the statement $S$ is executed in program state
$\state$, with continuation $\cont$, and with $\cost$ resources
available then---after a finite number of steps---the evaluation will
reach the new machine state $(\state', S', \cont',\cost')$ and there
are $\cost'$ resources available.  If $\cost' \ge 0$ then the
execution did not run out of resources and the resource consumption up
to this point is $\cost - \cost'$.  If this difference is negative
then resource became available during the execution.  If however
$\cost' < 0$ then the execution ran out of resources and is stuck.
The cost of the execution is then $\cost \ge 0$.

\pref{lem:sem1} summarizes the main properties of the resource
counter of the semantics.  It can be proved by induction on the number
of steps.
\begin{lemma}
\label{lem:sem1}
  \begin{enumerate}
  \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
    \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
    \smallstep{M}^k (\state', S', \cont', \cost' {+} n)$.
  \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
    \cont', \cost')$ and $(\state, S, \cont, d) \smallstep{M}^k (\state', S',
    \cont', d')$ then $\cost - \cost' = d - d'$.
  \end{enumerate}
\end{lemma}
%
If we assume that we have a \emph{positive resource metric}, that is,
a metric that assigns a positive cost to each evaluation step then we
can link bounds on the resource consumption under the metric to
termination.  This is formalized by \pref{lem:sem_term}.
%
\begin{lemma}
 \label{lem:sem_term}
 Let $M$ be a positive metric and assume that there exists $b \in
 \Qplus$ such that $(\state, S, \cont, \cost) \smallstep{M}^*
 (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.
 Then there exists $B \in \N$ such that $(\state, S, \cont, d)
 \smallstep{M}^n (\state'', S', \cont', d')$ for some $n$, $d$, $d'$
 and $\state''$ implies $n \le B$.
\end{lemma}
